Analysis

### 2023-02-10 Analysis.rmd
update code and figures

colors <- RColorBrewer::brewer.pal(n = 10, name = "Set1")

@@

```{r}
forecasts <- fread("data/forecast-data.csv")
scores <- fread("data/all-scores-crowd-forecasts.csv") |>
    mutate(type_and_scale = factor(type_and_scale, 
                                 levels = c("Cases - natural", "Deaths - natural", 
                                            "Cases - sqrt", "Deaths - sqrt", 
                                            "Cases - log", "Deaths - log")))
weekly_truth_revised <- fread("data/weekly-truth.csv")
daily_truth_revised <- fread("data/daily-truth.csv")
weekly_truth <- fread("data/weekly-truth-original.csv")
daily_truth <- fread("data/daily-truth-original.csv")

@@

models <- forecasts$model |> unique()

@@

plot_df |>
  ggplot(aes(x = target_end_date)) +
  geom_line(aes(y = prediction, group = interaction(forecast_date, model), 
                color = "Individual participant"), 
            linewidth = .1) + 
  geom_line(data = filter(plot_df, model == "crowd-ensemble"), 
            aes(y = prediction, group = interaction(forecast_date, model), 
                color = "Crowd ensemble"),
            linewidth = .4) + 
  geom_line(aes(y = true_value),
            linewidth = .4, color = "black") + 
  geom_point(aes(y = true_value),
             size = .5, color = "black") + 
  scale_color_manual(values = c(colors[1], "grey70")) + 
  facet_wrap(~ target_type, scale = "free_y", ncol = 1) + 
  labs(x = "Date", y = "Median forecast and observed values", color = "Forecaster") + 
  theme_scoringutils()
```

@@

comparison_plot(scores |>
                  filter(horizon == 2), 
                models = c("crowd-direct", "EuroCOVIDhub-ensemble"))

@@

df <- scores |>
  filter(model %in% models_non_human, 
         scale %in% c("log", "natural")) |>
    summarise_scores(by = c("horizon", "model", 
                            "target_type", "scale", "type_and_scale")) |>
  rename("Dispersion" = dispersion, 
         "Overprediction" = overprediction, 
         "Underprediction" = underprediction, 
         Model = model) |>
    pivot_longer(names_to = "WIS components",
                 cols = c(`Dispersion`, `Overprediction`, `Underprediction`)) |>
    mutate(`WIS components` = fct_relevel(`WIS components`, 
                                          c("Dispersion", "Underprediction", 
                                            "Overprediction")))

@@

cov <- forecasts |>
  filter(model %in% models_non_human) |>
  score() |>
  add_coverage(by = c("model", "target_type", "horizon")) |>
  summarise_scores(by = c("model", "target_type", "range", "horizon"))

cov_plot <- function(cov_scores, filter_range) {
  cov_scores |>
    filter(range == filter_range) |>
    ggplot(aes(x = horizon, y = coverage, color = model)) + 
    geom_line(show.legend = FALSE) + 
    geom_point(show.legend = FALSE) + 
    geom_hline(yintercept = filter_range / 100, 
               linetype = "dashed", colour = "grey80") + 
    facet_wrap(~ target_type, ncol = 2) + 
    scale_y_continuous(labels = \(x) {paste0(100 * x, "%")}) + 
    theme_scoringutils() + 
    labs(x = "Horizon", y = paste("Coverage -", filter_range, "PI"))
}

p_cov50 <- cov_plot(cov, 50)
p_cov90 <- cov_plot(cov, 90)

@@

p_performance

@@

@@

```{r, eval=FALSE}

coverage <- cov |>
  select(model, target_type, horizon, coverage_50, coverage_90) |>
  unique()

scores |>
  filter(model %in% models_non_human, 
         scale %in% c("log", "natural")) |>
  summarise_scores(by = c("model", "target_type", "scale", "horizon")) |>
  inner_join(coverage) |>
  filter(horizon == 2) |>
  select(model, target_type, scale, interval_score, coverage_50, coverage_90) |>
  pivot_wider(names_from = scale, values_from = interval_score) |>
  select(model, target_type, natural, log, coverage_50, coverage_90) |>
  mutate(across(c(where(is.numeric)), signif, digits = 2)) |>
  kable(format = "latex", booktabs = TRUE, 
        align = c("l", "l", rep("c", 4)),
        linesep = c('', '', '', '\\addlinespace'),
        col.names = c("Model", "Target", "WIS", "WIS (log scale)", 
                      "Coverage 50%", "Coverage 90%")) |> 
    kable_styling(latex_options = c("scale_down", 
                                    "hold_position")) 
```

@@

@@

  forecasts |>
    filter(model %in% models_non_human, 
           horizon == 2) |>
    filter(quantile %in% c(0.025, 0.25, 0.5, 0.75, 0.975)) %>%
    dcast(... ~ quantile, value.var = "prediction") |>
    mutate(target_type = paste(target_type, "-", scale)) |>
    ggplot(aes(x = target_end_date)) + 
    geom_linerange(aes(ymin = `0.25`, ymax = `0.75`, color = model), 
                   size = 1.1,
                   alpha = 1,
                   position = position_dodge(width = 4.8), 
                   show.legend = FALSE) + 
    geom_line(data = truth |>
                filter(target_end_date >= "2021-05-01", 
                       target_end_date <= "2021-09-01"), 
              aes(y = true_value)) + 
    geom_point(data = truth |>
                 filter(target_end_date >= "2021-05-01", 
                        target_end_date <= "2021-09-01"), 
               aes(y = true_value), size = 0.7) + 
    theme_scoringutils() +
    facet_wrap(~ target_type, scales = "free", ncol = 2) + 
    scale_y_continuous(labels = label_fn) + 
    theme(legend.position = "bottom") + 
    labs(y = "Forecasts", x = "Date", colour = "Model")
}

@@

plot_wis_components <- function(scores) {
  scores |>
    filter(model %in% models_non_human, 
           horizon == 2) |>
    summarise_scores(by = c("target_end_date", "model", "type_and_scale")) |>
    select(-interval_score) |>
    # add dispersion value to produce accurate "stacked" barplot
    mutate(underprediction = underprediction + dispersion, 
           overprediction = overprediction + dispersion) |>
    data.table::melt(measure.vars = c(
      "overprediction",
      "underprediction",
      "dispersion"
    ),
    variable.name = "wis_component_name",
    value.name = "component_value"
    ) |>
    mutate(wis_component_name = stringr::str_to_title(wis_component_name)) |>
    mutate(wis_component_name = factor(
      wis_component_name, 
      levels = c("Dispersion", "Overprediction", "Underprediction"))) |>
    mutate(target_end_date = as.Date(target_end_date)) |>
    ggplot(aes(x = target_end_date, group = model)) +
    geom_col(
      position = "dodge",
      aes(y = component_value, 
          fill = model, 
          group = model, alpha = wis_component_name), 
      color = "black", 
      size = 0.1,
      width = 5.5
    ) +
    facet_wrap(~ type_and_scale, scales = "free_y") +
    theme_scoringutils() +
    scale_alpha_manual(values = c(1, 0.6, 0.1)) +
    scale_y_continuous(labels = label_fn) + 
    expand_limits(x = as.Date(c("2021-05-01", "2021-09-01"))) +
    labs(y = "WIS", x = "Target end date", 
         fill = "Forecaster", alpha = "WIS component") +
    theme(legend.position = "bottom", 
          legend.box="vertical", legend.margin = margin()) 
}

@@

@@

>>>>

colors <- RColorBrewer::brewer.pal(n = 9, name = "Set1")

@@

add_type_and_scale <- function(data) {
  data |>
    mutate(type_and_scale = paste(target_type, "-", scale)) |>
    mutate(type_and_scale = factor(type_and_scale, 
                                   levels = c("Cases - natural", "Deaths - natural", 
                                              "Cases - log", "Deaths - log")))
}

forecasts <- fread("data/forecast-data.csv")
scores <- fread("data/all-scores-crowd-forecasts.csv") |>
  add_type_and_scale()

forecasts_revised <- fread("data/forecast-data-revised.csv")    
scores_revised <- fread(here("data", "all-scores-crowd-forecasts-revised.csv")) |>
  add_type_and_scale()

weekly_truth_revised <- fread("data/weekly-truth-revised.csv")
daily_truth_revised <- fread("data/daily-truth-revised.csv")
weekly_truth <- fread("data/weekly-truth-original.csv")
daily_truth <- fread("data/daily-truth-original.csv")

@@

Participant forecasts 
```{r}
models <- forecasts$model |> unique()

@@

plot_participant_forecasts <- plot_df |>
  ggplot(aes(x = target_end_date)) +
  geom_line(aes(y = prediction, group = interaction(forecast_date, model), 
                color = "Individual participant"), 
            linewidth = .1) + 
  geom_line(data = filter(plot_df, model == "crowd-ensemble"), 
            aes(y = prediction, group = interaction(forecast_date, model), 
                color = "Crowd ensemble"),
            linewidth = .4) + 
  geom_line(aes(y = true_value),
            linewidth = .4, color = "black") + 
  geom_point(aes(y = true_value),
             size = .5, color = "black") + 
  scale_color_manual(values = c(colors[1], "grey70")) + 
  facet_wrap(~ target_type, scale = "free_y", ncol = 1) + 
  labs(x = "Date", y = "Median forecast and observed values", color = "Forecaster") + 
  theme_scoringutils()

ggsave("output/figures/participant-forecasts.png", plot = plot_participant_forecasts, 
       width = 7, height = 5)
```

Median forecasts ensemble
```{r}

@@

comparison_plot(scores, 
                models = c("crowd-direct", "EuroCOVIDhub-ensemble"))

@@

plot_performance <- function(scores) {
  df <- scores |>
    filter(scale %in% c("log", "natural")) |>
    summarise_scores(by = c("horizon", "model", 
                            "target_type", "scale", "type_and_scale")) |>
    rename("Dispersion" = dispersion, 
           "Overprediction" = overprediction, 
           "Underprediction" = underprediction, 
           Model = model) |>
    pivot_longer(names_to = "WIS components",
                 cols = c(`Dispersion`, `Overprediction`, `Underprediction`)) |>
    mutate(`WIS components` = fct_relevel(`WIS components`, 
                                          c("Dispersion", "Underprediction", 
                                            "Overprediction")))

@@

  cov_plot <- function(scores, filter_range) {
    scores |>
      filter(range == filter_range) |>
      summarise_scores(by = c("model", "target_type", "range", "horizon")) |>
      ggplot(aes(x = horizon, y = coverage, color = model)) + 
      geom_line(show.legend = FALSE) + 
      geom_point(show.legend = FALSE) + 
      geom_hline(yintercept = filter_range / 100, 
                 linetype = "dashed", colour = "grey80") + 
      facet_wrap(~ target_type, ncol = 2) + 
      scale_y_continuous(labels = \(x) {paste0(100 * x, "%")}) + 
      theme_scoringutils() + 
      labs(x = "Horizon", y = paste("Coverage -", filter_range, "PI"))
  }

  p_cov50 <- cov_plot(scores, 50)
  p_cov90 <- cov_plot(scores, 90)

@@

  return(p_performance)
}

p_performance <- plot_performance(
  filter(scores, model %in% models_non_human))

@@

```{r}
# helper function to score forecasts both on the natural and on the log scale
score_and_add_log <- function(forecasts) {
  forecasts |>
  mutate(scale = "natural") |>
  rbind(forecasts |>
          mutate(
            scale = "log", 
            true_value = log(true_value + 1), 
            prediction = log(pmax(prediction, 0) + 1)
          )) |>
  score(metrics = c("interval_score", "coverage")) |>
  add_coverage(by = c("target_type", "horizon", "method")) |>
    add_type_and_scale() |>
  as.data.table() 
}

# helper function that filters the forecast data such that we only have 
# forecasts left where a single person made a forecast both for direct and rt
filter_comparison_rt_direct <- function(forecasts) {
  forecasts_comparison <- forecasts |>
  filter(model %in% models_human) |>
  mutate(rt_name = ifelse(
    grepl("(Rt)", model), 
    gsub("\ \\(Rt\\)", "", model), 
    NA
  )) |>
  mutate(method = ifelse(grepl("(Rt)", model), "rt", "direct")) |>
  group_by(forecast_date, target_type) |>
  filter(model %in% rt_name | rt_name %in% model) |>
  group_by(forecast_date, target_type, method) |>
  select(target_end_date, target_type, forecast_date, horizon, model, method, true_value, prediction, quantile)
}

forecasts_comparison <- filter_comparison_rt_direct(forecasts)
scores_forecasts_comparison <- score_and_add_log(forecasts_comparison)

forecasts_comparison_revised <- filter_comparison_rt_direct(forecasts_revised)
scores_forecasts_comparison_revised <- score_and_add_log(forecasts_comparison_revised)

scores_forecasts_comparison |> 
  filter(horizon == 2) |>
  mutate(name = gsub("\ \\(Rt\\)", "", model)) |>
  summarise_scores(by = c("target_type", "type_and_scale", "name", "method", "scale")) |>
  ggplot(aes(y = interval_score, x = name, group = method, color = method)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~ type_and_scale, scale = "free_y") + 
  theme_scoringutils()

plot_rt_direct_individual <- function(scores) {
 scores |> 
  filter(horizon == 2) |>
  mutate(name = gsub("\ \\(Rt\\)", "", model)) |>
  summarise_scores(by = c("target_type","type_and_scale", "name", "method", "scale")) |>
  arrange(target_type, name, method) |>
  group_by(target_type, name, scale) |>
  mutate(diff = rep(diff(interval_score), 2)) |>
  mutate(improvement = ifelse(diff > 0, "Direct better", "Rt better")) |>
  mutate(method = stringr::str_to_title(method)) |>
  ggplot(aes(y = interval_score, x = method, group = name)) + 
  geom_point(size = 0.5) + 
  geom_line(aes(color = improvement), linewidth = 0.3) + 
  facet_wrap(~ type_and_scale, scale = "free_y") + 
  labs(y = "WIS", x = "Forecasting method", color = "Comparison") + 
  scale_y_continuous(labels = label_fn) + 
  theme_scoringutils()
}

p_comparison_direct_rt_individual <- plot_rt_direct_individual(scores_forecasts_comparison)

ggsave("output/figures/comparison-direct-rt-individual.png", 
       plot = p_comparison_direct_rt_individual, 
       width = 7, height = 5)

p_comparison_direct_rt_individual_reivsed <- plot_rt_direct_individual(scores_forecasts_comparison_revised)

ggsave("output/figures/comparison-direct-rt-individual-revised.png", 
       plot = p_comparison_direct_rt_individual, 
       width = 7, height = 5)

# performance plot for an ensemble of the subset of human and direct forecasts
forecasts_comparable_ensemble <- forecasts_comparison |>
  group_by(target_type, horizon, method, quantile, true_value, forecast_date, target_end_date) |>
  summarise(prediction = median(prediction)) |>
  mutate(model = str_to_title(method)) |>
  as.data.table()

scores_comparable_ensemble <- score_and_add_log(forecasts_comparable_ensemble)

p_performance_comparison <- plot_performance(
  scores_comparable_ensemble
)

ggsave("output/figures/performance-rt-comparison.png", plot = p_performance_comparison, 
       width = 7, height = 8)

# performance plot for an ensemble of the subset of human and direct forecasts
# using revised death data

forecasts_comparable_ensemble_revised <- forecasts_comparison_revised |>
  group_by(target_type, horizon, method, quantile, true_value, forecast_date, target_end_date) |>
  summarise(prediction = median(prediction)) |>
  mutate(model = str_to_title(method)) |>
  as.data.table()

scores_comparable_ensemble_revised <- score_and_add_log(forecasts_comparable_ensemble_revised)

p_performance_comparison_revised <- plot_performance(
  scores_comparable_ensemble_revised
)

ggsave("output/figures/performance-rt-comparison-revised.png", plot = p_performance_comparison, 
       width = 7, height = 8)

scores_comparable_ensemble |>
  filter(horizon == 2) |>
  summarise_scores(by = c("target_type", "horizon", "model", "scale"))

scores |>
  filter(scale != "sqrt") |>
  filter(model %in% models_non_human) |>
  filter(horizon == 2,
         scale == "log") |>
  summarise_scores(by = c("target_type", "horizon", "model", "scale"))

@@

```{r, performance-table}
make_performance_table <- function(scores) {
  scores |>
    filter(scale %in% c("log", "natural")) |>
    summarise_scores(by = c("model", "target_type", "scale", "horizon")) |>
    filter(horizon == 2) |>
    select(model, target_type, scale, interval_score, coverage_50, coverage_90) |>
    pivot_wider(names_from = scale, values_from = interval_score) |>
    select(model, target_type, natural, log, coverage_50, coverage_90) |>
    mutate(across(c(where(is.numeric)), signif, digits = 2)) |>
    kable(format = "latex", booktabs = TRUE, 
          align = c("l", "l", rep("c", 4)),
          caption = "Performance for two-week-ahead forecasts \\label{tab:scores}",
          linesep = c('', '', '', '\\addlinespace'),
          col.names = c("Model", "Target", "WIS", "WIS (log scale)", 
                        "Coverage 50%", "Coverage 90%")) |> 
    kable_styling(latex_options = c("scale_down", 
                                    "hold_position")) 
}

scores |>
  filter(model %in% models_non_human) |>
  make_performance_table() |>
  cat(file = 'manuscript/performance-table.tex')

scores_revised |>
  filter(model %in% models_non_human) |>
  make_performance_table() |>
  cat(file = 'manuscript/performance-table-revised.tex')

scores_comparable_ensemble |>
  make_performance_table() |>
  cat(file = 'manuscript/performance-table-comparison-rt.tex')

@@

plot_scores_and_pred <- function(forecasts, scores) {

@@

    forecasts |>
      filter(horizon == 2) |>
      filter(quantile %in% c(0.025, 0.25, 0.5, 0.75, 0.975)) %>%
      pivot_wider(values_from = prediction, names_from = quantile) |>
      # dcast(... ~ quantile, value.var = "prediction") |>
      mutate(target_type = paste(target_type, "-", scale)) |>
      ggplot(aes(x = target_end_date)) + 
      geom_linerange(aes(ymin = `0.25`, ymax = `0.75`, color = model), 
                     size = 1.1,
                     alpha = 1,
                     position = position_dodge(width = 4.8), 
                     show.legend = FALSE) + 
      geom_line(data = truth |>
                  filter(target_end_date >= "2021-05-01", 
                         target_end_date <= "2021-09-01"), 
                aes(y = true_value)) + 
      geom_point(data = truth |>
                   filter(target_end_date >= "2021-05-01", 
                          target_end_date <= "2021-09-01"), 
                 aes(y = true_value), size = 0.7) + 
      theme_scoringutils() +
      facet_wrap(~ target_type, scales = "free", ncol = 2) + 
      scale_y_continuous(labels = label_fn) + 
      theme(legend.position = "bottom") + 
      labs(y = "Forecasts", x = "Date", colour = "Model")
  }

@@

  plot_wis_components <- function(scores) {
    scores |>
      filter(horizon == 2) |>
      summarise_scores(by = c("target_end_date", "model", "type_and_scale")) |>
      select(-interval_score) |>
      # add dispersion value to produce accurate "stacked" barplot
      mutate(underprediction = underprediction + dispersion, 
             overprediction = overprediction + dispersion) |>
      data.table::melt(measure.vars = c(
        "overprediction",
        "underprediction",
        "dispersion"
      ),
      variable.name = "wis_component_name",
      value.name = "component_value"
      ) |>
      mutate(wis_component_name = stringr::str_to_title(wis_component_name)) |>
      mutate(wis_component_name = factor(
        wis_component_name, 
        levels = c("Dispersion", "Overprediction", "Underprediction"))) |>
      mutate(target_end_date = as.Date(target_end_date)) |>
      ggplot(aes(x = target_end_date, group = model)) +
      geom_col(
        position = "dodge",
        aes(y = component_value, 
            fill = model, 
            group = model, alpha = wis_component_name), 
        color = "black", 
        size = 0.1,
        width = 5.5
      ) +
      facet_wrap(~ type_and_scale, scales = "free_y") +
      theme_scoringutils() +
      scale_alpha_manual(values = c(1, 0.6, 0.1)) +
      scale_y_continuous(labels = label_fn) + 
      expand_limits(x = as.Date(c("2021-05-01", "2021-09-01"))) +
      labs(y = "WIS", x = "Target end date", 
           fill = "Forecaster", alpha = "WIS component") +
      theme(legend.position = "bottom", 
            legend.box="vertical", legend.margin = margin()) 
  }

@@

  return(p_scores_and_pred)  
}

p_scores_and_pred <- plot_scores_and_pred(
  filter(forecasts, model %in% models_non_human),
  filter(scores, model %in% models_non_human)
)

@@

p_scores_and_pred_comparison <- plot_scores_and_pred(
  forecasts_comparable_ensemble,
  scores_comparable_ensemble
)

ggsave("output/figures/scores-and-forecasts-rt-comparison.png", 
       plot = p_scores_and_pred_comparison, 
       width = 9, height = 9)
